---
title: "R Basics Continued"
output:
        html_document:
            includes:
                in_header: header.html
            theme: paper
            toc: true
            number_sections: false
            fig_caption: false
            markdown: GFM
            code_download: false
---

<style type="text/css">

body, td {
   font-size: 18px;
}
code.r{
  font-size: 12px;
}
pre {
  font-size: 12px
}

</style>

<!--
---
title: "R Basics continued - factors and data frames"
teaching: 60
exercises: 30
questions:
- "How do I get started with tabular data (e.g. spreadsheets) in R?"
- "What are some best practices for reading data into R?"
- "How do I save tabular data generated in R?"
objectives:
- "Explain the basic principle of tidy datasets"
- "Be able to load a tabular dataset using base R functions"
- "Be able to determine the structure of a data frame including its dimensions and the datatypes of variables"
- "Be able to subset/retrieve values from a data frame"
- "Understand how R may coerce data into different modes"
- "Be able to change the mode of an object"
- "Understand that R uses factors to store and manipulate categorical data"
- "Be able to manipulate a factor, including subsetting and reordering"
- "Be able to apply an arithmetic function to a data frame"
- "Be able to coerce the class of an object (including variables in a data frame)"
- "Be able to import data from Excel"
- "Be able to save a data frame as a delimited file"
keypoints:
- "It is easy to import data into R from tabular formats including Excel. However, you still need to check that R has imported and interpreted your data correctly"
- "There are best practices for organizing your data (keeping it tidy) and R is great for this"
- "Base R has many useful functions for manipulating your data, but all of R's capabilities are greatly enhanced by software packages developed by the community"
source: Rmd
---
-->

```{r chunk_options, include = FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("03-")
```

# Overview

In this lesson we will learn how to access, subset, and otherwise operate on the contents of the objects we encountered in the previous lesson.

# Accessing and subsetting objects

In the last lesson we created the following objects:

```{r}
countries
country_populations
continents
continents_factor
keep_data
country_list
country_matrix
country_df
```

But how do we access the individual elements of these objects? Depending on the type, the method of access may be different, and for the same type, there can be alternative methods of access. As always, the best way to see this is with some examples.

## Atomic vectors

Atomic vectors have a notion of length, and are indexed starting from 1. Some languages index positions starting from 0 (Python, for example), but R starts from 1. We can see the length of an atomic vector with the `length()` function:

```{r}
length(countries)
```

To access an element, or elements, of an atomic vector, we use single-bracket notation, as in:

```{r}
countries[1]
countries[3]
country_populations[2]
continents[4]
```

We can also subset atomic vectors by passing a vector of indices within the square brackets, as in:

```{r}
# Extract indices 1 through 3
countries[1:3]
# Extract indices 1, 3, and 4
country_populations[c(1,3,4)]
```

A `logical` vector can also be passed into the square brackets, so long as it has the same length as the vector we are subsetting. Recall our logical vector has the values:

```{r}
keep_data
```

What do we think will happen if we put this in the square brackets of the `countries` vector?

```{r}
countries[keep_data]
```

R interprets the `TRUE` values in the logical vector as indices to keep, and `FALSE` as indices to drop. So Suriname, the third element of `countries` is not displayed. Using logical vectors to subset data is often a powerful way to programatically subset data, as we will see in future lessons.

## Lists

Accessing elements in lists is a little more complex in that the method used determines whether a subsetted list or an atomic vector is returned. Let's investigate what happens with some examples. Just as the atomic vector `countries` had a length, so does `country_list`:

```{r}
country_list
length(country_list)
```

To access the first element of `country_list` and **return a list**, we would use the single-bracket notation:

```{r}
country_list[1]
typeof(country_list[1])
```

If we wanted to access the first element and **return an atomic vector**, we would use double-bracket notation:

```{r}
country_list[[1]]
typeof(country_list[[1]])
```

Right now, appreciating this subtlety is not crucial. However, you may discover in your future experience that you wish to operate on the elements of lists. Then returning the correct type via the correct mode of access will be important. To see this with a quick example, say we wanted to divide all the populations by one million. We would first need to access the `population` element of `country_list`, but which mode of access should we use? Let's try both, specifying the named `population` element instead of it's numerical index:

```{r error = TRUE}
# Double-bracket notation returns an atomic vector
country_list[['populations']] / 1000000
# Single-bracket notation returns a list
country_list['populations'] / 1000000
```

In the first case, R understands how to divide numbers by a number; the double-bracket subsetting returns a vector of numbers. In the second case, R doesn't know how to divide a list by a number, and so it throws an error.

There is another way to access elements of a list by name, and that is with dollar-sign notation. This method will **only return the atomic vector**.

```{r}
# Dollar-sign notation always returns an atomic vector
country_list$countries
```

We may need to access individual elements within the elements of lists. For example the following four commands all return the first country in the `countries` element of `country_list`, Thailand:

```{r}
# Access with numerical indices
country_list[[1]][1]
# Access with mixture of named and numerical indices
country_list[['countries']][1]
# Access with mixture of named and numerical indices and all double-bracket notation
country_list[['countries']][[1]]
# Access with mixture of dollar-sign named index and numerical index
country_list$countries[1]
```

If we want to access multiple elements of a list, we use single-bracket notation, as in:

```{r}
# Access multiple list entries by index
country_list[1:2]

# Access multiple list entries by their names, if they exist
country_list[c('countries','continents')]

# Access multiple list entries with a logical vector
country_list[c(TRUE, FALSE, TRUE)]
```

It is illustrative to try the above commands using double-bracket notation:

```{r error = TRUE, purl = FALSE}
country_list[[1:2]]
country_list[[c('countries','continents')]]
```

The first form returned just a single country, which is perhaps unexpected. It returned the first element of `country_list`, and the second element of that atomic vector. However, the second form returned an error, because double-bracket notation returns a vector.

We have gone a bit into the weeds here, but hopefully this exposition will set off a reminder in your brain when accessing elements of lists in the future.

## Matrices

As we saw earlier, matrices have two-dimensions. We can get the dimensions of a matrix with the `dim()` function:

```{r}
country_matrix
dim(country_matrix)
```

The result is first the number of rows, then the number of columns. This suggests that when we want to access elements of a matrix, we need to specify two pieces of information, the row index and the column index. We can also use the `nrow()` and `ncol()` functions to get the dimensions individually:

```{r}
nrow(country_matrix)
ncol(country_matrix)
```

To access elements of the matrix, we use single-bracket notation, as in:

```{r}
# Return the element in the first row, second column
country_matrix[1, 2]

# Return the element in the second row, second column
country_matrix[2, 2]
```

If we wanted to return a whole row, or a whole column, we would leave the row or column entry blank, respectively, as in:

```{r}
# Return the first row
country_matrix[1, ]

# Return the second column
country_matrix[, 2]
```



## Data Frames

Remember that data frames group vectors together into a two-dimensional table. As with matrices, we can use the `dim()`, `nrow()`, and `ncol()` functions to see the number of rows and columns of a data frame.

```{r}
country_df
dim(country_df)
nrow(country_df)
ncol(country_df)
```

To access elements, rows, or columns of a data frame, we use similar methods to matrices and lists that we saw earlier. To access a single element, we can use row and column indices in single-bracket notation as in:

```{r}
# Return the element in the second row, third column
country_df[2, 3]
```

Since the rows and columns of the data frame are named, we can also use those instead:

```{r}
# Return the element in the Lesotho row of the population column
country_df['Lesotho', 'populations']
```

We can access rows in the same way as matrices:

```{r}
# Access the third row by index
country_df[3, ]
# Or by name
country_df['Suriname', ]
```

And it is the same with columns:

```{r}
# Access the second column by index
country_df[, 2]
# By name
country_df[, 'continents']
# Or by name with dollar sign notation
country_df$continents
```

Note, the dollar-sign notation only accesses columns, it cannot be used to access rows. We can also access multiple rows, as with:

```{r}
# Second and third rows
country_df[2:3, ]
country_df[c('Lesotho', 'Suriname'), ]
```

Or columns with:

```{r}
country_df[, c(1,3)]
country_df[, c('countries','populations')]
```
And of course, these can be combined to subset the data frame into another data frame, as in:

 ```{r}
country_df[c('Thailand', 'Canada'), c('countries', 'populations')]
 ```

Finally, as with vectors, we can also subset rows or columns of data frames with logical vectors. Let's use the `keep_data` vector to subset the rows of `country_df`:

```{r}
country_df[keep_data, ]
```

> ### A Warning

Could we use `keep_data` to subset the columns of `country_df`? Why or why not?

```{r error = TRUE}
country_df[, keep_data]
```

Turns out no, the `keep_data` vector has length 4, but the `country_df` only has 3 columns. But let's look at one more example with logical subsetting to make ourselves aware of behavior we may not expect, or want.

```{r}
country_df[, c(TRUE, FALSE)]
```

What has R done here? Since `country_df` has 3 columns, and the logical vector we're using to subset the columns only has length 2, R will **repeat** the vector until it reaches the number of columns. So it's really doing:

```{r}
country_df[, c(TRUE, FALSE, TRUE)]
```

This illustrates that code can do unexpected things, and these unexpected things might cause errors in analysis, or downstream errors that are hard to detect.

In the above examples we input a logical directly to subset rows or columns, but we can also use a conditional statement. For example, if we wanted to subset the rows of `country_df` to only those countries with more than 10,000,000 inhabitants, we could do:

```{r}
country_df[country_df$populations > 10000000, ]
```

What has R done here? It first looked at the code inside of the square brackets and evaluated it to:

```{r}
country_df$populations > 10000000
```

And it used this logical vector in the same way it did when we manually defined a vector of `TRUE` and `FALSE` values.

In all the code above, we did not assign the result of any of our commands to a new variable, so the accessing/subsetting was not persistent--it was not stored. If we wanted to create a new data frame that only shows Thailand's and Canada's populations, we could create a new variable and assign it the correct value:

```{r}
subset_df <- country_df[c('Thailand', 'Canada'), c('countries', 'populations')]
subset_df
```

# Summary

In this lesson we learned how to access and subset vectors, lists, matrices, and data frames using numerical indices, named indices, and logical vectors (sometimes in the form of conditionals). Understanding the different data types we can have, and how to manipulate them is a crucial skill before we can analyze and plot data.
